---
title: "Seminar 8 - Displaying Data with Dashboards"
author: "Dr Nikolaos Korfiatis"
date: "Week 8"

output: 
  pdf_document: 
    toc: yes
    number_sections: yes
header-includes:
    - \usepackage{hyperref}
always_allow_html: true
---
\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Shiny 

* Shiny is the dashboard component of R and is developed by RStudio as a viable
container for displaying results and interacting with the user 

* For this seminar we work with the ui.R and server.R files in this folder. 

* To begin - go to File > New File > Shiny Web App

## Structure of a shiny app

A shiny app consists of two main vars: 

* ui: Handling the interface and user interaction 
* server: Handling the processing of the data on the environment

For each of them we assign properties and relevant functions

## The shinydashboard package 

* It adds visual appealing using known frameworks such as Twitter Bootstrap 
which allow for the quick design of dashboard visualizations

* It consists of three parts 
  * header (dashboardHeader())
  * sidebar (dashboardSidebar())
  * body (dashboardBody())
* Note: All these need to be declared on the ui.R file.
  
# Declaring a basic dashboard 

* We edit the ui.R file in order to create a new 
dashboardPage and assign it to the ui component

```{r dashboard, eval=FALSE}
library(shinydashboard)

ui <- dashboardPage(
  dashboardHeader(title = "Basic dashboard"),
  dashboardSidebar(),
  dashboardBody()
)
```


## Adding elements to the dashboard page 

* Shinydashboard follows twitter bootstrap 

* Every element is positioned in rows and columns using the 
fluidRow() and column() functions. 

* Elements of interest are provided in boxes using the box() function


## Example: Creating our hello world 

* Add a row with two boxes: One box has the world hello and the 
other has the word World

* To do that we need to split the row in two sections using the 
column function. Our structure looks as follows: 


```{r helloworld, eval=FALSE}
fluidRow(
  column(width = 6,
         box(p("hello"))
         ), 
  column(width = 6, 
         box(p("world"))
  )
)


```

## HTML tags and formating

* Notice we use the p() function to display a paragraph. 
This is the html equivalent of 
```<p>Hello</p><p>World</p>```

* The column functions allow us to position the boxes side by side 

* We can change the p to h1(), h2(), h3() header functions. 


## Adding structure on the sidebar 

* We can beggin by adding a basic structure 
* For this example we are going to use the diamonds dataset from 
ggplot 
* We want to have a page where we list the data and two additional 
sections where we provide relevant visualizations 

## Sidebar items 

* Lets begin with the sidebar panel

* We use the menuItem() function with the appropriate parameters 
for label, reference to action and icon 

```{r sidebarcontent, eval=FALSE}
## Sidebar content
  dashboardSidebar(
       sidebarMenu(
         menuItem("Data", tabName = "diamondsdata", icon = icon("dashboard")),
         menuItem("Plots", tabName = "diamondsplots", icon = icon("th"))
    )
  )
```


## Adding actions for sidebar item 

* We can now add separate action for each dashboard menu item on 
the dashboardBody page 

* Lets start with two empty pages 
```{r dashboardbody, eval=FALSE}
dashboardBody(
    tabItems(
      # First tab content
      tabItem(tabName = "diamondsdata",
        fluidRow(
          h1("Dataset")
        )
         
      ),

      # Second tab content
      tabItem(tabName = "diamondsplots",
        h1("Plots")
      )
    )
  )

```

# Displaying data using the DT package 

```{r dtrender, eval=FALSE}
# add this on the libraries start 
library(DT)

# lets add a fluidRow with the dataTableOutput 
# function from DT package

DT::dataTableOutput("diamondsdatatable")

```

## Connecting the datatable with the server.R 

* Now that we have placed a component to be displayed lets modify the server.R 
to do that 

* Lets open server.R 

* replace the contents of this file with the following: 

```{r server,eval=FALSE}
server <- function(input,output){
    
}
```

* Now we need to connect the component on the ui.R with the server.R functions 

## DataTables rendering 

* Remember that in the ui.R we have now added an output function with the parameter
argument: ``diamondsdatatable``

* To connect now, we have to use the output component from the server function and 
use the renderDataTable() function to fill up the dataTableOutput() function. 

* We can do this as follows: 
```{r display, eval=FALSE}
output$diamondsdatatable <- DT::renderDataTable(
        diamonds
    )
```

## Wrapping up 

* We can also add the datatable within a column to make the interface more usable 

```{r shinycolumn, eval=FALSE}
fluidRow(
          column(width = 12,
                  DT::dataTableOutput("diamondsdatatable")
          )
)
```


# Displaying visuals 

* We achieve that using the plotOutput function 
* Note that this type of function evaluates an expression, meaning the that the server part will be as: 

```{r renderplot, eval=FALSE}
 output$pricehist <- renderPlot({
        ggplot(diamonds, aes(y=price,x=carat))+geom_smooth()
        }
    )
```

* We can also use the plotly library to create a more interactive set of graphs

```{r plotout, eval=FALSE}

    output$pricehistplotly <- renderPlotly({
        output <-  ggplot(diamonds, aes(y=price,x=carat))+geom_smooth()
        ggplotly(output)
      }
    )
```

* Note that you need to modify the ``plotOutput`` to ``plotlyOutput``


## Adding controls 

* We can add a selectbox 

```{r seelect, eval=FALSE}
selectInput("diamondcut", label = h3("Select Cut"), 
                                         choices = list("Fair" = "Fair", "Good"="Good", 
                                                        "Very Good"="Very Good", 
                                                        "Premium"="Premium", 
                                                        "Ideal"="Ideal" 
                                                        ), 
                                         selected = "Ideal")
```

## Connecting the filter 
```{r plotfilter, eval=FALSE}
output$pricehistplotlywithfilter <- renderPlotly({
        to_filter <- input$diamondcut
        ploth <- ggplot(subset(diamonds,cut==to_filter),aes(x=price))+geom_histogram()
        ggplotly(ploth)
    })
```



## Finishing up and upload 

* You can use the publish option to upload it to shinyapps.io 
* Note: You need to create an account there and use the token to publish it 
through Rstudio.



# Datacamp course 

* Building Dashboards with Shiny dashboard 

https://learn.datacamp.com/courses/building-dashboards-with-shinydashboard