---
title: 'Seminar 5 - Advanced SELECT and JOINS'
author: "Dr Nikolaos Korfiatis"
date: "Week 5"
# output:
#   beamer_presentation:
#     theme: Boadilla
#     colortheme: dolphin
#     fonttheme: structurebold
#     slide_level: 2
# classoption: t
output: 
  pdf_document: 
    toc: yes
    number_sections: yes
header-includes:
    - \usepackage{hyperref}

---
\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,comment=NA,attr.source='.numberLines')

rm(list=ls())
library(readr)
library(RSQLite)
library(dplyr)
```


# Preliminaries 

* We are going to work with the bibitor database we created in the previous 
  seminar 



```{r connect}

my_db <- RSQLite::dbConnect(RSQLite::SQLite(),"bibitor.db")

```

# Advanced Select 

## Aliasing 

```{sql connection=my_db}
-- Usefull to know, we can actually get the list of tables 
-- from the metadata in the database using sql and no need 
-- to actually go back to the terminal
--SELECT type, name, rootpage 
--FROM sqlite_master 
--ORDER BY rootpage

SELECT name,type
FROM sqlite_master 
WHERE type ='table'
```
* Let us remember the VendorInvoices table and how it looks 


```{sql connection=my_db}
SELECT * 
FROM VendorInvoicesDec
```

### Calculating complex sums 

>From table ‘VendorInvoicesDec ’ obtain the total freight costs by VendorNumber for
transactions which the Total Dollars value of the
transaction was greater than $100 and the quantity
was less than or equal to 1000 units. 

>Which VendorNumber corresponded to the largest Freight
cost under these conditions?


```{sql connection=my_db}
SELECT VendorNumber,VendorName,SUM(Freight) AS total_freight_costs, SUM(Freight)/SUM(Quantity) as avg_cost_per_unit
FROM VendorInvoicesDec
WHERE Dollars > 100 AND Quantity <= 1000
GROUP BY VendorNumber
HAVING total_freight_costs >1000

ORDER BY avg_cost_per_unit DESC
--LIMIT 1
```



### Using aliases 

```{sql connection=my_db}
SELECT DISTINCT VendorNumber, VendorName 
FROM VendorInvoicesDec
```

This will give us the same results as the query above. Notice that the column 
names are not altered

```{sql connection=my_db}
SELECT DISTINCT p.VendorNumber, p.VendorName 
FROM VendorInvoicesDec p
```


## LIKE 

* The like statement is used to appropriate where conditions using pattern matching 
* It is only applicable on non-numeric columns 
* The general syntax of like is as follows: 

```{sql, eval=FALSE}
SELECT colname1,colname2,.... 
FROM TABLE 
WHERE colname1 LIKE '%endswith' OR 
WHERE colname2 LIKE 'startswith%' OR 
WHERE colname3 LIKE '%contains%' 
```

* Lets use it to select all suppliers with the name baccardi 

```{sql connection=my_db}
SELECT DISTINCT VendorName 
FROM VendorInvoicesDec
WHERE VendorName LIKE '%' 
```

* Notice that the LIKE statement is not case sensitive. That means that we 
can conveniently use upper case or lower case characters depending on the 
context.

* Which vendors in this table have the USA in their name ? 


```{sql connection=my_db}
SELECT DISTINCT VendorName 
FROM VendorInvoicesDec
WHERE VendorName LIKE '%corp%'
```


## UNION 

* So far we have seen how we can get information from separate queries in 
  different result sets. 
* Using the union operator we can combine them together in one 

```{sql connection=my_db}
SELECT COUNT(*) AS value, 'Total Invoices' AS metric
FROM VendorInvoicesDec 
UNION 

SELECT SUM(Dollars) AS value, 'Total Ammount' AS metric 
FROM VendorInvoicesDec

UNION 

SELECT MIN(Dollars) AS value, 'Minimum Ammount' AS metric 
FROM VendorInvoicesDec


UNION 

SELECT SUM(Freight) AS value, 'Total weight' AS metric
FROM VendorInvoicesDec
```
* Observe that the UNION is meaningful only if we specify the same 
amount of columns in all queries and have the same name.

* Run the same query with a different set of columns


## Views and temporary result storage 

* A VIEW is a virtual table. That means that the results are available 
for select but not for insert/update/delete. 
* The basic commands CREATE and DROP from SQL DDL apply to Views as well. 
* Contrary to tables Views require 

```{sql connection=my_db}
DROP VIEW IF EXISTS BACCARDI
```


```{sql connection=my_db}
CREATE VIEW IF NOT EXISTS BACCARDI AS 
SELECT * 
FROM VendorInvoicesDec 
WHERE VendorName = "BACARDI USA INC"	
```



* We can add the IF EXISTS operator here to avoid getting an error message
if we do it again 

```{sql connection=my_db}
SELECT * 
FROM BACCARDI
```

* if a new record is added for baccardi, will the view table contain that ? 


## Imposing relations using the WHERE statement 

* Let us consider the two tables: ``BegInvDec`` and ``EndInvDec`` 
* They contain the same number of columns however in the ``BegInvDec`` we have the ``startDate`` and in the ``EndInvDec`` we have the ``endDate`` 

* Lets now select ``InventoryId`` from ``BeginvDec`` with the start date 
and the ``endDate`` from the ``EndInvDec`` 

```{sql connection=my_db}
SELECT a.InventoryId, a.Brand, a.startDate, b.endDate, b.endDate - a.startDate AS total_days_inv
FROM BegInvDec a, EndInvDec b 
WHERE a.InventoryId = b.InventoryId
ORDER BY total_days_inv DESC
```

* Can we get the vendor name and merge it so we can have a look ? 
* Lets create a view for that (remove it if exists)

```{sql connection=my_db}
DROP VIEW IF EXISTS brand_vendor
```

```{sql connection=my_db}
CREATE VIEW brand_vendor AS 
SELECT DISTINCT Brand,VendorNumber,VendorName 
FROM PurchasesDec
```

* Lets expand now (takes a bit of time)

> Note that views can be treated as tables

```{sql connection=my_db}
SELECT a.InventoryId,a.Brand, a.startDate, b.endDate, b.endDate-a.startDate AS total_days_inv, c.VendorNumber, c.VendorName
FROM BegInvDec a, EndInvDec b, brand_vendor c
WHERE a.InventoryId = b.InventoryId AND a.Brand = c.Brand
ORDER BY total_days_inv DESC
LIMIT 5
```



# JOINS 

## Inner Join 

* The characteristic of the Inner Join is the matching of one column with a column on another table. 
* In other words it works similar with a where statement 

* Lets check the original record count

```{sql connection=my_db}
SELECT COUNT(*) as original_count
FROM BegInvDec
```


```{sql connection=my_db}
-- Count the inventory that went unsold at the end of the year  
-- meaning that it was recorded on the beggining and still available 
-- at the end of the year 
-- requires the cross check between the two tables
SELECT COUNT(*) AS total_unsold
FROM EndInvDec A 
INNER JOIN BegInvDec B 
ON A.InventoryId = B.InventoryId

```


## Outer(Left) Join 

* Outer join is commonly known by the direction of the join 
* LEFT join includes all records from Table A and those from 
Table B that can be matched with the join key of Table A
* RIGHT join includes all records from Table B and those from Table A that can be matched with the join key of Table B 

```{sql connection=my_db}

SELECT a.InventoryId,a.Brand, a.startDate, b.endDate, b.endDate-a.startDate AS total_days_inv, c.VendorNumber, c.VendorName
FROM BegInvDec a 
LEFT JOIN EndInvDec b ON a.InventoryId = b.InventoryId 
LEFT JOIN PurchasesDec c ON a.Brand = c.Brand
ORDER BY total_days_inv DESC
-- add a limit as the result set will be long 
LIMIT 5 
```

* Similar cases can be done when considering the right join.


## Cross Join 

* Cross join is meaningfull if we want to get all possible 
combinations of table A with table B. 

* This is particularly usefull when we work with combined fields 
such as date. 

* Lets consider the following example: 

  * year_tbl: a set of years from 2000 to 2020 
  * month_tbl: the months within a year from January to February 
* Lets write these tables in the database so we can play with 

```{r addcrossjoin}
years_tbl <- data.frame(years = c(2000:2021))
months_tbl <- data.frame(months = month.name)

# write them to the db so we execute an sql statement 
dbWriteTable(my_db,"years_tbl",years_tbl,overwrite =TRUE)
dbWriteTable(my_db,"months_tbl",months_tbl,overwrite =TRUE)
```

```{sql connection=my_db}
SELECT * FROM years_tbl 
CROSS JOIN months_tbl
```



# Dataset  

* We continue with the second database on the bibitor case - the one 
on the bibitor2 folder.


## Sequential loading 

* Lets consider the contents of the files in the second folder (bibitor2)
* These contain the purchases per month
* Let us load the first file (2016PurchasePricescsv.zip)

## Unzipping

* Copy the folder bibitor2 from Week 4 

```{bash,eval=FALSE}
# Tip we can use unzip "*.zip" to get 
# the originals out 
cd bibitor2 
unzip "*.zip"
```

## Let sniff the data

```{r readzip}
purchase_prices <- read.csv("bibitor2/2016PurchasePricesDeca.csv")
```
* Ok it seems that the dataset has the same dimension accross all the files 

## Merging the different files together

* Lets create a dataframe here where we get them all together 

```{r files, message=FALSE,warning=FALSE}

# create an empty dataframe 
purchase_prices <- data.frame()

filesh <- list.files("bibitor2/",pattern = "csv")
```

```{r loopfor,message=FALSE,warning=FALSE}
for(i in 1:length(filesh)){
  print(paste0("reading for file: ",filesh[i]))
  this_file <- readr::read_csv(paste0("bibitor2/",filesh[i]))
  this_file$filename <- filesh[i]
  purchase_prices <- rbind(purchase_prices,this_file)
}
```

## Add year and quarter information 

* Lets create a dataframe to join with the case we have here

```{r getfilenames}

yearquarter <- data.frame(filename =unique(purchase_prices$filename))
```

```{r extractyear}
# first four digits is the year 
yearquarter$year <- substr(yearquarter$filename,1,4)
```

* Quarter information is released as follows: 

March : Q1 
June  : Q2 
Sep   : Q3 
Dec   : Q4 


```{r extractquarter}
# easy to understand approach 
# to a check for the string of the filename for every row  
# not very efficient for very large dataframes though 

#let us first create a new column to keep the quarterly 
#information 
yearquarter$quarter <- NA

for(i in 1:nrow(yearquarter)){
 # Set a variable that changes on every iteration 
 quarterh <- NA
 #check the value of the filename 
 if(grepl("Mar",yearquarter$filename[i])){
    quarterh <- "Q1"
 }
 else if(grepl("Jun",yearquarter$filename[i])){
    quarterh <- "Q2"
 }
 else if(grepl("Sep",yearquarter$filename[i])){
    quarterh <- "Q3"
 }
 else if(grepl("Dec",yearquarter$filename[i])){
    quarterh <- "Q4"
 }
 # now lets set the variable to the corresponding cell position 
 yearquarter$quarter[i] <- quarterh

 
 print(quarterh)
}

```


## Merge the file together 

```{r mergejoin}
library(dplyr)

purchase_prices %>% 
  left_join(yearquarter) -> purchase_prices

# remove also the filename column since we got the information 
# we wanted 

purchase_prices$filename <- NULL 
```


## Advanced group by using dplyr

```{r purprices}
purchase_prices %>% 
  mutate(profit = Price - PurchasePrice) %>% 
  group_by(Brand,year,quarter) %>% 
  summarise(avprofit = mean(profit)) 
  
```
* Since we have the information for quarter and year lets  plot the price datapoints 
* We are going to use both pipes and expression connectors (+) using ggplot

```{r funplotting}
#load the library as well

library(ggplot2)


purchase_prices %>% 
  group_by(year,quarter) %>% 
  summarise(avgprice = median(Price)) %>% 
  ggplot(.,aes(x=paste0(year,"-",quarter),y=avgprice))+
  geom_bar(stat="identity")+
  ylab("Average Price")+
  xlab("Year (Quarter)")+
  # ggthemes::theme_wsj()+
  theme(axis.text.x = element_text(angle = 90))+
  ggtitle("Prices go up :)")


```


* Has the price of our stock increased accross these four years ? 

# Recommended Datacamp courses. 

* Datacamp course Joining data in sql 
https://learn.datacamp.com/courses/joining-data-in-postgresql

* Datacamp course on Joining data with dplyr 
https://learn.datacamp.com/courses/joining-data-with-dplyr
