---
title: "Seminar 1 - Flat files, Data Structures and DB Interfaces"
author: "Dr Nikolaos Korfiatis"
date: "Week 1"
output: 
  pdf_document: 
    toc: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview 
The goal of this seminar is to familiarize you with: 

* File system commands and flat files
* Data reading  
* Use of the DBI api from R to connect to SQLite 


## File System commands 

To learn more about the commands we are going to use see the related courses 
from Datacamp on the category shell

* The ls command - this command is just listing the files in the current working
directory

```{bash}
ls
```

* To get help you can use the --help option 

```{bash, comment=NA}
ls --help
```

* The --help is a *flag*, it is used as an input to the command, in a similar way
we may call a function in R with a named argument. There are several alternative 
flags that we may want to utilize by running the ``ls --help`` command again 
and seeing what is useful 

* lets get a more detailed view of the ls commnand using the ``ls -l`` command

```{bash}
ls -l
```

* The defualt block size is 4096, total is counting the number of disk blocks. 

* The letter ``d`` in front of the second line denotes that ``challenge`` is a folder.

* We can use even more flags to accomplish what we want. For example let us see the size of theses files in a human readable way. By using the 



```{bash, comment=NA}
ls -lahsr
```

The ``list.files()`` function in R performs the same job and returns a vector.

* Sequential peeking into a file  

The head command 

```{bash, comment=NA}
cd challenge/
head textfile.txt
```

* The cat command 

```{bash}
#go to the challenge folder
cd challenge 
#print out the contents of the text file 
cat textfile.txt

```

We can also use the ``-n`` flag to show the line number 

```{bash,comment=NA}
#go to the challenge folder
cd challenge 
#print out the contents of the text file 
cat -n textfile.txt

```


## Working with Flat files


```{bash,comment=NA}
cd challenge
head data.csv
```

* We can use the pipe (|) symbol to channel the output of the cat command to another 
command - ``wc``. This command counts the number of lines provided in the input and 
provides an output. We can use the ``-l`` 


```{bash,comment=NA}
cd challenge
# how many lines
cat data.csv | wc -l

```

```{bash,comment=NA}
#contents of specific lines using sed
#let us just grab line number: 1,999,222
cd challenge
sed -n -e 1999222p data.csv

```

* We can also request multiple rows by adding the -e flag sequentially 

```{bash,comment=NA}
#contents of specific lines using sed
#let us  grab line number: 1,999,222 and 
#add line 1 which is the header row.
cd challenge
sed -n -e 1p -e 1999222p data.csv

```

## The RSQLite package

```{r package,comment=NA}
# install.packages("RSQLite")
library(RSQLite)
```

Setting up the connection 

```{r connection,eval=FALSE}
#declare the connection to a file called test.db 
#alternatively if you don't want the file to be created and reside in the memory 
#you can use the ":memory:" option 

#First we open the connection to the database called test.db 
#if the database doesn't exist a file will be created
dbconnection <- dbConnect(RSQLite::SQLite(), "test.db")

#Lets see how many tables are there 
dbListTables(dbconnection)

#Let us write a table called mtcars 
dbWriteTable(dbconnection, "mtcars", mtcars)

#If an error message says that the table already exists we can 
#add the overwrite or append option as follows: 
# dbWriteTable(dbconnection, "mtcars", mtcars,overwrite=TRUE)


#Do not forget - after we do anything we disconnect
dbDisconnect(dbconnection)
```






