---
title: 'Seminar 4 - SQL SELECT and dplyr'
author: "Dr Nikolaos Korfiatis"
date: "Week 4"
# output:
#   beamer_presentation:
#     theme: Boadilla
#     colortheme: dolphin
#     fonttheme: structurebold
#     slide_level: 2
# classoption: t
output: 
  pdf_document: 
    toc: yes
    number_sections: yes
header-includes:
    - \usepackage{hyperref}

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,comment=NA,attr.source='.numberLines')

rm(list=ls())
library(readr)
library(RSQLite)
library(dplyr)
```

\newpage

# Dataset


## The bibitor data 

* The bibitor data are part of the hub for analytics education PWC developed 
  interview case study / training 
* They rely on two folders for this seminar: bibitor and bibitor2 


## Loading the data in SQLite 

* The case study considers how to load the data using SQLite browser
* We are going to load the data using the methods taught in Seminar Week 2 


```{r defineconnection}
# we will use the name of this variable to 
# set up sql output in the rmd file

my_connection <- RSQLite::dbConnect(RSQLite::SQLite(),"bibitor.db")
```

* We are going to read the bibitor data as dataframes using the names of dataframes 
as the SQL tables defined in the PWC case study 

## Loading using ``read_csv`` and ``dbWriteTable()``

```{r dataloading,message=FALSE,warning=FALSE}
PurchasesDec <- readr::read_csv("bibitor/PurchasesFINAL12312016a.csv")
BegInvDec <- readr::read_csv("bibitor/BegInvFINAL12312016a.csv")
EndInvDec <- readr::read_csv("bibitor/EndInvFINAL12312016a.csv")

# in a similar way we can read and load the other tables and give the names  
# that are in the PWC data case study - see slide 9 from the uploaded pdf
VendorInvoicesDec <- readr::read_csv("bibitor/InvoicePurchases12312016a.csv",show_col_types = FALSE)
```

## Write them to the database 
```{r writebacktodb}
RSQLite::dbWriteTable(my_connection,"PurchasesDec",PurchasesDec,overwrite=TRUE)
RSQLite::dbWriteTable(my_connection,"BegInvDec",BegInvDec,overwrite=TRUE)
RSQLite::dbWriteTable(my_connection,"EndInvDec",EndInvDec,overwrite=TRUE)
RSQLite::dbWriteTable(my_connection,"VendorInvoicesDec",VendorInvoicesDec,overwrite=TRUE)

```

# SQL Select 


## General Syntax 

* The general syntax is as follows (``|`` denotes ``or``, ``[]`` denotes non 
mandatory): 

```{sql, eval=FALSE}
SELECT colname|expresion [AS alias] 
FROM tablename|other tables
WHERE condition 
GROUP BY colname
HAVING condition
ORDER BY colname [ASC|DESC]
LIMIT [start]TotalRecords
```

## Examples 
>tip: We can use the rmarkdown to connect directly to the SQLite without 
having to go to the sqlite console


### Selecting columns 

* Get the VendorName
```{sql connection=my_connection}
SELECT VendorName
FROM VendorInvoicesDec ;
```



* Get the VendorName and other columns

```{sql connection=my_connection}
SELECT VendorNumber, VendorName,Dollars 
FROM VendorInvoicesDec ;
```


## The wildcard (``*``) expression 

We can also select all columns using the star (*) or wildcard expression 


```{sql connection=my_connection}
SELECT * 
FROM VendorInvoicesDec 
```


## The WHERE statement 


```{sql connection=my_connection}
SELECT * 
FROM VendorInvoicesDec 
WHERE VendorName = "MARTIGNETTI COMPANIES" 
```



```{sql connection=my_connection}
SELECT * 
FROM VendorInvoicesDec 
WHERE upper(VendorName) = upper("Martignetti COMPANIES") 
```

* Remember condition value is case sensitive 



* Same with multiple columns 

```{sql connection=my_connection}
SELECT * 
FROM VendorInvoicesDec 
WHERE VendorName="MARTIGNETTI COMPANIES" OR VendorName="BACARDI USA INC"
```

* What will happen if we use AND instead of OR when using the same column  ? 

```{sql connection=my_connection}
SELECT * FROM VendorInvoicesDec 
WHERE VendorName="MARTIGNETTI COMPANIES" AND VendorName="BACARDI USA INC" 
```


* The IN operator 

```{sql connection=my_connection}
SELECT * 
FROM VendorInvoicesDec 
WHERE VendorName IN("MARTIGNETTI COMPANIES","BACARDI USA INC") 
```



* Example with numbers 

```{sql connection=my_connection}
SELECT * FROM VendorInvoicesDec 
WHERE PONumber >= 8150 
```

## WHERE using multiple conditions

* Example with numbers 
```{sql connection=my_connection}
SELECT * 
FROM VendorInvoicesDec 
WHERE PONumber >= 8150 AND PONumber < 8170 
```

* Same example using the between keyword 

* Example with numbers , observe the difference in the record numbers returned 
* Between uses the equal sing in addition
```{sql connection=my_connection}
SELECT * 
FROM VendorInvoicesDec 
WHERE PONumber BETWEEN 8150 AND 8170
```

## Using the LIMIT  

* The limit allows us to set limits of the recordset to be returned by the query
planner. 
* The syntax of LIMIT considers the start row and the extend of the rows to be 
returned e.g., LIMIT 10 will return the first 10 rows. LiMIT 10,10 will return 
the ten rows after row 10 etc. 


```{sql connection=my_connection}
SELECT * 
FROM VendorInvoicesDec 
LIMIT 5
```

* Largest invoice (order by clause)

```{sql connection=my_connection}
SELECT VendorName, Quantity
FROM VendorInvoicesDec 
ORDER BY Dollars DESC
LIMIT 5 
```


## The DISTINCT condition 

* The distinct condition considers the cases where we get unique instances 
of the values in that column 
* It always consider column names 

```{sql connection=my_connection}
SELECT DISTINCT VendorName FROM VendorInvoicesDec 
```

## Summarization Functions 

Summarization functions in SQL are the following: 

  * AVG – calculates the average of a set of values.
  * COUNT – counts rows in a specified table or view.
  * MIN – gets the minimum value in a set of values.
  * MAX – gets the maximum value in a set of values.
  * SUM – calculates the sum of values.


## Summarization on the whole table 

* Counting the number of rows in the table

```{sql connection=my_connection}
SELECT COUNT(*) 
FROM VendorInvoicesDec;
```

using the AS to make the resulted column conform with the standards 

```{sql connection=my_connection}
SELECT COUNT(*) AS total_rows FROM VendorInvoicesDec;
```


* The number of distinct suppliers that we have represented 

```{sql connection=my_connection}
SELECT COUNT(DISTINCT VendorName) AS total_vendors
FROM VendorInvoicesDec;
```

* The minimum value of an invoice 

```{sql connection=my_connection}
SELECT MIN(Dollars) AS min_value FROM VendorInvoicesDec;
```

* The minimum value of an invoice in euros, assuming 1 euro is 0.82 dollars 


```{sql connection=my_connection}
SELECT MIN(Dollars*0.86) AS min_value_euros FROM VendorInvoicesDec;
```


* The maximum value of an invoice 

```{sql connection=my_connection}
SELECT MAX(Dollars) AS max_value FROM VendorInvoicesDec;
```

* The total value of all invoices 

```{sql connection=my_connection}
SELECT SUM(Dollars) AS total_value 
FROM VendorInvoicesDec;
```

## Summarisation and Grouping 

* Very often as in the case of pivot tables we want breakdowns by a particular 
category 

* This is achieved with the GROUP BY statement followed by the column or columns 
where we want the pivot table to be disected. 

* Lets get the total payment per vendor 
 - Expand the original
 
```{sql connection=my_connection}
SELECT SUM(Dollars) AS total_value 
FROM VendorInvoicesDec 
GROUP BY VendorName 
```

```{sql connection=my_connection}
SELECT VendorName, SUM(Dollars) AS total_payment 
FROM VendorInvoicesDec 
GROUP BY VendorName 
```

* It will make more sense if we get the value in desc order instead of 
an ascending 

```{sql connection=my_connection}
SELECT VendorName, SUM(Dollars) AS total_payment 
FROM VendorInvoicesDec 
GROUP BY VendorName 
ORDER BY total_payment DESC
```

* We can also add the LIMIT 5 to get the top 5 vendors by costs of goods bought

```{sql connection=my_connection}
SELECT VendorName, SUM(Dollars) AS total_payment 
FROM VendorInvoicesDec 
GROUP BY VendorName 
ORDER BY total_payment DESC 
LIMIT 5;
```


* Instead of the LIMIT I can use the having to get the suppliers that have cost 
of goods purchased more than 1 million

```{sql connection=my_connection}
SELECT  VendorName, SUM(Dollars) AS total_payment 
FROM VendorInvoicesDec 
GROUP BY VendorName 
HAVING total_payment > 1000000
ORDER BY total_payment DESC 
```


## Running the SQL directly to the database and bringing it to R 

* We use the dbGetQuery method 

```{r top5vendors}

top_suppliers <- RSQLite::dbGetQuery(my_connection,"SELECT VendorName, 
                                            SUM(Dollars) AS total_payment
                                            FROM VendorInvoicesDec 
                                            GROUP BY VendorName 
                                            ORDER BY total_payment DESC")

```

```{r barplot}
barplot(top_suppliers$total_payment)
```



```{r disconnection}
RSQLite::dbDisconnect(my_connection)
```



# Dplyr 


## An example of the previous query

* SQL 

```{sql, eval=FALSE}
SELECT VendorName, SUM(Dollars) AS total_payment
FROM VendorInvoicesDec 
GROUP BY VendorName 
ORDER BY total_payment DESC 
LIMIT 5
```


* dplyr example

```{r dplyr}
VendorInvoicesDec %>%
  group_by(VendorName) %>% 
  summarise(total_payment = sum(Dollars)) %>% 
  arrange(desc(total_payment)) %>% 
  top_n(5)
```
## Working with another table : BegInvDec


```{r bibitor}
head(BegInvDec)
```


* The following fragment is the dplyr equivalent of the following sql command 
```{sql, eval=FALSE}
SELECT Price,PPrice, Price*0.82 AS price_euros,
PPrice*0.82 AS purchase_price_euros
FROM BegInvDec
LIMIT 5
```


```{r dplyravg}

BegInvDec %>%                    
  select(Price,PPrice) %>%         
  mutate(price_euros = Price*0.82,                       
        purchase_price_euros = PPrice*0.82) 

```

# Recommended Data Camp courses for Practicing 

* Introduction to SQL 
https://learn.datacamp.com/courses/introduction-to-sql

* Introduction to data manipulation with dplyr 
https://learn.datacamp.com /courses/data-manipulation-with-dplyr