---
title: "Seminar 7 - Crawling Data"
author: "Dr Nikolaos Korfiatis"
date: "Week 7"
output:
  pdf_document:
    toc: yes
    number_sections: yes
  word_document:
    toc: yes
header-includes: \usepackage{hyperref}
always_allow_html: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
rm(list=ls())
library(dplyr)
library(rvest)
library(tidyr)
library(stringr)
library(ggplot2)
```

\newpage


# Pagination and Reviews 

* We want to get the average rating and the number of reviews for all banks in trustpilot 

* Lets fetch the following page 
> https://uk.trustpilot.com/categories/bank


```{r trustpilot}
trustpilot_url <- "https://uk.trustpilot.com/categories/bank?page=1"
```


```{r readpagination}
trustpilot_pages <- read_html(trustpilot_url)
```

* From this page we want to find the urls of all bank pages in trustpilot
* We need to find the title of the bank the number of the reviews and the rating
selector

```{r names}
trustpilot_pages %>% 
  html_nodes(".styles_businessTitle__1IANo") %>% 
  html_text() -> bank_names 

bank_names

```

```{r ratings}
trustpilot_pages %>% 
  html_nodes(".styles_textRating__19_fv") %>% 
  html_text() -> bank_reviews
bank_reviews
```


```{r individualpageurls}
trustpilot_pages %>% 
  html_node(".styles_categoryBusinessListWrapper__2H2X5") %>% 
  html_nodes(".link_internal__YpiJI") %>% 
  html_attr('href') -> individual_urls
individual_urls
```

* merge 

```{r singlepage}
all_banks_trustpilot <- tibble(
  bank_name = bank_names, 
  bank_review = bank_reviews,
  individual_url = individual_urls
)

```


* do some cleanup 

```{r cleanup}
library(tidyr)

all_banks_trustpilot <- all_banks_trustpilot %>% 
  tidyr::separate(bank_review, into=c("reviews","rating_score"),sep="·")
```

```{r review_count_rating}
all_banks_trustpilot$rating_score <- gsub("TrustScore ","",all_banks_trustpilot$rating_score)
all_banks_trustpilot$rating_score <- stringr::str_trim(all_banks_trustpilot$rating_score)
# convert to numeric 
all_banks_trustpilot$rating_score <- as.double(all_banks_trustpilot$rating_score)


```

* How do ratings look ? 

```{r histplot}
library(ggplot2)
all_banks_trustpilot %>% 
  ggplot(.,aes(x=rating_score))+geom_histogram()
```

* Do the same for the number of reviews

```{r reviewcount}
all_banks_trustpilot$reviews <- gsub("reviews","",all_banks_trustpilot$reviews)
all_banks_trustpilot$reviews <- gsub(",","",all_banks_trustpilot$reviews)
all_banks_trustpilot$reviews <- stringr::str_trim(all_banks_trustpilot$reviews)
all_banks_trustpilot$reviews <- as.numeric(all_banks_trustpilot$reviews)
```

* Review count and rating 

```{r ratingreviewcount}

all_banks_trustpilot %>% 
  ggplot(.,aes(x=reviews,y=rating_score))+geom_smooth()
```
* Why do we see review score stabilizing after a certain amount of reviews ? 

# S&P and Russel Index  


* the html_table() function allows us to extract the data that are in tabular 
form avoiding for loops


* get the companies on the russel index

```{r getrusselindex}
russel_index <- "https://en.wikipedia.org/wiki/Russell_1000_Index"
read_html(russel_index) %>% 
  html_table() %>% 
  .[[3]] -> russel_index_companies 

```

* get the companies on the S&P index 

```{r sandpindex}
spindex <- "https://en.wikipedia.org/wiki/List_of_S%26P_500_companies"
read_html(spindex) %>% 
  html_table(fill = TRUE) %>% 
  .[[1]] -> sp500_companies

```


* lets get an inner join of the companies that appear both at the russel index 
and the S&P500

```{r sprussel}
# fix the column names so we have a common field 
colnames(russel_index_companies) <- c("company","symbol")
colnames(sp500_companies) <- tolower(colnames(sp500_companies))
# remove also the spaces and dashes from the sp500 dataframe column names 
colnames(sp500_companies) <- gsub(" ","_",colnames(sp500_companies))
colnames(sp500_companies) <- gsub("-","_",colnames(sp500_companies))
```

```{r innerjoin}
sp500_companies %>% 
  inner_join(russel_index_companies)
```



```{r antijoin}
#we can also do an anti join to get those in S&P 500 not in Russel 1000 
sp500_companies %>% 
  anti_join(russel_index_companies)
```


```{r keepall}
sp500_companies %>% 
  anti_join(russel_index_companies) %>% 
  select(symbol,security,cik) 
```


```{r geturlsof10k}
edgar_url <- "https://www.sec.gov/cgi-bin/browse-edgar?CIK=%s&type=10-k"

tenK_urls <- data.frame()
#lets do the first 6 
for(i in 1:nrow(head(sp500_companies))){
  # use the sprintf function instead of paste when we have long 
  # strings 
  url_to_fetch <- sprintf(edgar_url,sp500_companies$cik[i])
  this_row <- tibble(
    cik = sp500_companies$cik[i],
    url_to_fetch = url_to_fetch
  )
  tenK_urls <- bind_rows(tenK_urls,this_row)
  print(url_to_fetch)
}

```

* The following should run locally 

```{r navigate10kurls, eval=FALSE}
#initialize a column of the most recent document to NA
tenK_urls$document_url <- NA 

for(r in 1:nrow(tenK_urls)){
  print(paste("Fetching for CIK code:",tenK_urls$cik[r]))
  this_page <- read_html(tenK_urls$url_to_fetch[r])
  # from this page we want the first row of the table  
  # which corresponds to the more recent 10k filing of the company
  # using the selector we are now able to see that the .tableFile2 is the selector we 
  # wish to use to access the page
  this_page %>% 
    html_node(".tableFile2") %>% 
    html_nodes("tr") %>%
    .[[2]] %>% 
    html_node("a") %>% 
    html_attr("href") -> document_url
  
  tenK_urls$document_url[r] <- document_url
}
```

```{r fixdownload}
# use the paste0 function to add the URL 
tenK_urls$document_url <- paste0("https://www.sec.gov",tenK_urls$document_url)

```

```{r downloadfiles}
tenK_urls$document_url
```


